---
title: "Extending msuiter: third-party engines and the registry contract"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extending msuiter: third-party engines and the registry contract}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(msuiter)
```

This vignette is the extension contract promised by ARCHITECTURE.md section 3.4:
it walks through everything a third-party package must do to add a new
extraction engine to msuiter, and everything it promises in return. The example
engine uses nothing but base R and msuiter itself, so you can copy the whole
tutorial into your own package and adapt it.

The registry is *data, not a class explosion*: one row per engine name, keyed
by the engine name and carrying the spec class, the implementation function and
three governance fields (`engine_version`, `contract_version`, `certified`).
Your engine becomes available to every user of `ms_extract(method =)` and
`method = "<name>"` string sugar the moment its row exists.

## 1. The engine spec: an `MsEngine` subclass

Engine specs are small, serializable S7 objects. Create a subclass of the
abstract `MsEngine` with `S7::new_class()` and give its properties *defaults
that form a valid spec* — string sugar (`method = "tutorial_eu"`) resolves by
constructing the class with no arguments and checking the result against the
registry key.

```{r engine-class}
TutorialEu <- S7::new_class(
  "TutorialEu", package = "tutorial.othersig", parent = MsEngine,
  properties = list(
    name          = S7::new_property(S7::class_character, default = "tutorial_eu"),
    mode          = S7::new_property(S7::class_character, default = "extract"),
    deterministic = S7::new_property(S7::class_logical,   default = TRUE),
    packages      = S7::new_property(S7::class_character, default = character(0)),
    label         = S7::new_property(S7::class_character,
                                     default = "Tutorial Frobenius NMF (base R)"),
    max_iter      = S7::new_property(S7::class_numeric,   default = 200),
    seed          = S7::new_property(S7::class_numeric,   default = 1)
  )
)
```

Contract notes on the properties:

* `name`, `mode`, `deterministic` **must** have valid defaults: string sugar
  constructs the class bare, and `match_ms_engine()` rejects classes whose
  defaults fail validation or disagree with the registry key.
* `seed` is required for `mode = "extract"` engines: the `ms_extract()`
  assembly records the spec's `seed` on the returned `MsSignature` for
  provenance. A spec class without a `seed` property does not stop the
  assembly: it degrades to a registry warning and the `seed = 1` default —
  declare the property anyway so your provenance is real, even if your
  algorithm is fully deterministic.
* `deterministic = TRUE` is a promise: same inputs, same output, independent
  of thread count. The engine below owns its RNG draws (seeded) and is pure R,
  so the promise is easy to keep.
* Hyper-parameters (`max_iter`, `seed` here) are *frozen on the spec object*
  at construction time — the object, not a config file, is the unit of
  reproducibility.

## 2. The fit function

The registry calls your implementation as `fit_fn(engine, catalog = , k = )`
through `fit_engine()`. For an `"extract"` engine it must return a plain list
with two **raw, unnormalized** matrices:

* `signatures`: `channels x k`
* `exposures`: `k x samples`

Assembly into a labelled `MsSignature` (display normalization, labels, the
catalog snapshot) is `ms_extract()`'s job, not yours. The layout contract is
strict: channels in rows, in the catalog's channel order.

```{r fit-function}
tutorial_eu_fit <- function(engine, catalog, k, ...) {
  # Error protocol: violations raise msuiter_error_<topic> through
  # rlang::abort, carrying the i/j/c information payload (ARCHITECTURE.md
  # section 3.5). Never call stop().
  if (!S7::S7_inherits(catalog, MsCatalog)) {
    rlang::abort(
      "catalog must be an MsCatalog object",
      class = "msuiter_error_input",
      msuiter_error = list(
        topic = "input",
        i = "the tutorial engine factorizes a channel x sample count matrix",
        j = paste0("received: ", class(catalog)[1L]),
        c = "tally variants with ms_tally(), or build one with ms_catalog()"
      )
    )
  }
  if (!is.numeric(k) || length(k) != 1L || is.na(k) || k < 1 || k != trunc(k)) {
    rlang::abort(
      "k must be a single positive whole number",
      class = "msuiter_error_input",
      msuiter_error = list(
        topic = "input",
        i = "a rank-k factorization needs one signature per rank",
        j = paste0("received: ", format(k)),
        c = "pass e.g. k = 3"
      )
    )
  }
  k <- as.integer(k)
  counts <- catalog@counts # channels x samples, non-negative

  # Determinism: seed our own draws, and leave the caller's RNG state alone.
  had_seed <- exists(".Random.seed", envir = globalenv())
  if (had_seed) old_seed <- get(".Random.seed", envir = globalenv())
  on.exit(
    if (had_seed) assign(".Random.seed", old_seed, envir = globalenv()),
    add = TRUE
  )
  set.seed(engine@seed)

  m <- nrow(counts)
  n <- ncol(counts)
  eps <- 1e-12
  # Initialization: a uniform share for every channel plus one contiguous
  # anchor block per signature (seeded jitter breaks ties). Plain
  # multiplicative updates then converge without collapsing a component.
  block <- rep(seq_len(k), each = ceiling(m / k))[seq_len(m)]
  W <- matrix(1 / m, m, k)
  W[cbind(seq_len(m), block)] <- 1 / k
  W <- W * (1 + 0.1 * stats::runif(m * k))
  W <- sweep(W, 2, colSums(W), "/") # gauge: columns sum to 1
  H <- matrix(mean(counts), k, n) * (0.5 + stats::runif(k * n))
  # Lee-Seung multiplicative updates, Frobenius ("eu") divergence: the
  # denominator couples both factors, so the scale of W and H cannot run
  # away. Numerical guards live only in the divisions.
  for (iter in seq_len(engine@max_iter)) {
    H <- H * (crossprod(W, counts) + eps) / (crossprod(W, W %*% H) + eps)
    W <- W * (tcrossprod(counts, H) + eps) / (W %*% tcrossprod(H, H) + eps)
  }
  list(signatures = W, exposures = H) # raw factors: m x k and k x n
}
```

Anything your function raises as an `msuiter_error_*` condition propagates to
the caller as-is: `ms_extract()` wraps dispatch, not your errors. Raise them
with the three-part payload (`i` = what invariant holds, `j` = what arrived,
`c` = what to do) so users and tests can match on
`cnd$msuiter_error` (`topic`/`i`/`j`/`c`).

## 3. Register the engine

```{r register}
register_ms_engine(
  name = "tutorial_eu",
  mode = "extract",
  engine_class = TutorialEu,
  fit_fn = tutorial_eu_fit,
  packages = character(0),
  tags = c("third-party", "tutorial", "nmf"),
  engine_version = "0.1.0",
  contract_version = "1",
  certified = "self-reported"
)
```

The nine fields, and what each one owes the registry:

| field | meaning |
|---|---|
| `name` | registry key; single lower-case snake_case string (`^[a-z][a-z0-9_]*$`), unique per session |
| `mode` | `"extract"` (signature discovery) or `"fit"` (reference-based fitting) |
| `engine_class` | S7 class deriving from `MsEngine` with valid property defaults |
| `fit_fn` | the implementation, called as `fit_fn(engine, catalog = , k = )` |
| `packages` | R packages your engine needs, reported by `required_pkgs()` |
| `tags` | free-form labels (e.g. for benchmark filtering) |
| `engine_version` | version of *your* engine implementation |
| `contract_version` | version of the msuiter registry/extension contract you coded against (`"1"` today) |
| `certified` | one of `"certified"`, `"self-reported"`, `"unknown"` — see governance below |

Duplicate names are rejected (registration errors instead of silently
shadowing), and `contract_version` is your pin: when msuiter changes the
extension contract, the version bumps and stale engines are visible in
`ms_engines()` at a glance.

```{r engines-table}
ms_engines()
```

## 4. Trust boundary and governance

`certified` is the trust level of the row:

* `"certified"` — the engine passed the msuiter maintainers' compatibility
  review. In-tree engines (`nmf`, `ard`, `sparse`) are registered this way.
* `"self-reported"` — the engine's author vouches for the contract. Third
  parties almost always land here.
* `"unknown"` — trust level not established.

Trust boundary, stated plainly (from the `register_ms_engine()` docs):
**registered engine code runs with your permissions.** Registering and
resolving an engine runs *its* class and function objects inside *your*
session, with your user's rights. Only attach engines from packages you
trust, exactly as you would trust any R package you install.

Non-certified engines are loud, not silent: the first resolution of a
`"self-reported"` or `"unknown"` engine in a session — through
`match_ms_engine()`, and therefore also through `ms_extract()` /
`ms_fit()` / any benchmark that resolves the engine — warns once per session
per engine with condition class `msuiter_warning_uncertified_engine`. The
warning fires where the engine enters the computation, so a benchmark that
consumes a non-certified engine carries the warning with it; certified
engines never warn.

## 5. End to end: extract signatures with the tutorial engine

A small synthetic catalog: 48 channels x 8 samples built from three planted
signatures with disjoint channel blocks (seeded, so the vignette is
reproducible). In a real analysis the catalog would come from
`ms_variants() |> ms_tally()`.

```{r synthetic-catalog}
set.seed(11)
ch <- 48L; ns <- 8L; k_true <- 3L
planted <- t(vapply(seq_len(k_true) - 1L, function(s) {
  v <- numeric(ch)
  v[(s * 16L + 1L):((s + 1L) * 16L)] <- 0.5 + stats::runif(16L)
  v / sum(v)
}, numeric(ch))) # k_true x channels, disjoint supports
H_true <- matrix(2, k_true, ns)
for (j in seq_len(ns)) H_true[(j - 1L) %% k_true + 1L, j] <- 60
counts <- round(100 * crossprod(planted, H_true))
dimnames(counts) <- list(paste0("C", seq_len(ch)), paste0("S", seq_len(ns)))

cat1 <- ms_catalog(
  counts = counts,
  channels = list(name = "SYN48", labels = paste0("C", seq_len(ch))),
  samples = paste0("S", seq_len(ns)),
  provenance = list(genome = "SYN")
)
```

String sugar first. Because the engine is `"self-reported"`, the first
resolution warns (once per session):

```{r extract-sugar}
fit <- ms_extract(cat1, k = 3, method = "tutorial_eu")
```

The spec object is the first-class citizen (A14): passing `TutorialEu()`
carries the frozen hyper-parameters and warns no further — the one-time
warning is idempotent per engine.

```{r extract-object}
fit2 <- ms_extract(cat1, 3, TutorialEu(max_iter = 500, seed = 7))
identical(fit2@engine, "tutorial_eu")
```

The result is a standard `MsSignature`, indistinguishable from one produced
by an in-tree engine: the engine name and seed ride along for provenance.

```{r signature}
fit
```

Recovery check against the planted truth — each planted signature against its
best-matching recovered signature (cosine):

```{r recovery}
recovered <- fit@signatures # channels x signatures, columns sum to 1
planted_n <- sweep(planted, 1, sqrt(rowSums(planted^2)), "/")
recovered_n <- sweep(recovered, 2, sqrt(colSums(recovered^2)), "/")
round(apply(planted_n %*% recovered_n, 1, max), 4)
```

The raw-factor contract is also visible directly: `fit_engine()` dispatches
to the registered `fit_fn`, and `required_pkgs()` reports the registry row.

```{r raw-contract}
spec <- match_ms_engine("tutorial_eu")
raw <- fit_engine(spec, catalog = cat1, k = 3)
dim(raw$signatures); dim(raw$exposures)
required_pkgs(spec)
```

## 6. Shipping the engine in your package

In a real package the registration belongs in `.onLoad()`, so the row exists
before any user call can resolve it:

```{r, eval = FALSE}
# yourpackage/R/zzz.R
.onLoad <- function(libname, pkgname) {
  msuiter::register_ms_engine(
    name = "othersig_kl", mode = "fit", engine_class = OthersigKlEngine,
    fit_fn = othersig_kl_fit, packages = "othersig",
    engine_version = "2.3.4", contract_version = "1",
    certified = "self-reported"
  )
}
```

Notes:

* The registry is **per-session state**: it ships empty and lives in memory,
  so your package must re-register on every load — which `.onLoad()` does.
* `S7::methods_register()` is *not* needed for registration itself (registry
  rows are data). Call it in your `.onLoad()` only if your package defines
  its own S7 methods on msuiter generics (say, a `format()` method for your
  spec class), and only after those definitions.
* Guard idempotently if you source the registration outside a package
  context; inside `.onLoad()` it runs once per session anyway.
* Pick a name namespaced to your package (`"othersig_kl"`, not `"kl"`) — the
  registry is shared by every attached package in the session, and duplicate
  keys are rejected.

## 7. Your obligations: error protocol and contract tests

**Error protocol** (ARCHITECTURE.md section 3.5). Every violation your engine
detects raises `msuiter_error_<topic>` via `rlang::abort()` with the
`msuiter_error` payload named `c("topic", "i", "j", "c")` — the three-part
information shown in section 2. Bare `stop()` is a contract violation:
callers match on the condition class and the payload, not on message text.

**Contract-test checklist** — what msuiter expects an engine (and its
package) to test, mirroring the in-tree suites
(`tests/testthat/test-engines-registry.R` and
`tests/testthat/test-extend-contract.R`, which you can read as executable
templates):

1. Registration stores all nine fields; `ms_engines()` lists the row.
2. Duplicate names are rejected and the existing row survives.
3. Field violations raise `msuiter_error_registry`: name grammar, unknown
   mode, non-`MsEngine` class, non-function `fit_fn`, `NA`/empty entries in
   `packages`/`tags`, and missing or invalid governance fields
   (`engine_version`, `contract_version`, `certified`).
4. `certified` accepts exactly the three governance states.
5. String sugar resolves; an unknown name fails with *every* available engine
   name listed in the message.
6. A spec object passes through `match_ms_engine()` unchanged.
7. `fit_engine()` dispatches to the registered `fit_fn` (dots pass through);
   an unregistered spec fails with `msuiter_error_registry`.
8. `required_pkgs()` prefers the registry row.
9. The `.onLoad()` registration path works end to end, and non-certified
   engines warn exactly once per session per engine (string path *and*
   object path).
10. An engine class without usable defaults fails resolution with a clear
    `msuiter_error_registry`.
11. End to end through `ms_extract()`: your engine produces an `MsSignature`
    with the right `engine` name, `seed`, dimensions and labels; a degenerate
    fit (a signature column without positive mass) surfaces as
    `msuiter_error_signature`; engine failures surface with their own
    `msuiter_error_*` class; `"fit"`-mode engines are rejected by
    `ms_extract()` with `msuiter_error_input`.

Items 1-10 are registry-side and identical for every engine; item 11 is yours
to reproduce for your implementation.

## 8. Session notes

* `certified = "certified"` is reserved for the maintainer review (A17);
  self-certifying as `"certified"` is a contract violation. The maintainers
  re-run a monthly reverse-dependency compatibility matrix over certified
  engines; `"self-reported"` engines are revisited on maintainer request.
* There is no deregistration API: to clear the registry (tests, exploration),
  restart the session. Package-internal test helpers use
  `msuiter:::msuiter_registry_reset()`.
* The contract this vignette documents is `contract_version = "1"`. Read the
  `contract_version` field on every `ms_engines()` row before relying on
  cross-engine metadata.
```